\documentclass{article}
\title{ELEE 204 - Digital Logic Design\\Group Project}
\date{}
\usepackage[T1]{fontenc}
\usepackage{arev}
\usepackage{array}
\usepackage[hmargin=2cm,vmargin=2cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead{}
\fancyfoot{}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[LO,LE]{ELEE204}
\fancyhead[CO,CE]{Digital Logic Design}
\fancyhead[RO,RE]{Spring 2009}
\cfoot{\thepage}
\begin{document}
\maketitle
\thispagestyle{fancy}
Working in groups of 2-4, solve a problem using digital logic that has at least 6 inputs and at least 4 outputs.  The problem must require some kind of state information (memory), meaning it needs sequential circuits.  The design must use at least 8 total components of at least 4 different types that we learned about in class, such as decoders, MUXes, adders, etc.  For example, you might have 4 MUXes, 2 decoders, 1 encoder and 1 adder.  Most likely, this should be a problem that calls for an embedded system to make some decision or control something in its environment.
\\ \\
Document the problem in an objective way.  Describe your design goals, assigning an appropriate priority to each (i.e. some goals are more important than others).  Design the necessary circuit on paper.  Then implement the design on a breadboard, on the Altera board or on some type of PLD.  Perform tests to verify the design matches the problem and goals.  Document your design process, key decisions, rationale for those decisions and your evaluation.
\\ \\
It would probably help you if you talked to me after coming up with a preliminary problem statement and a rough idea of the design approach.
\\ \\
Specifically, I would like to turn in the following documention:
\begin{itemize}
  \item A prioritized list of design goals
  \item A block diagram
  \item A circuit diagram
\end{itemize}
\textbf{Important Dates}
\begin{itemize}
  \item Have one member of your group email me the names of the members of your group by Tuesday, March 31.
  \item Turn in your design documentation on Tuesday, April 21.
  \item You'll need to present the finshed, working circuit in class on Tuesday, May 5.  Please prepare a brief (less then 5 minute) presentation which includes the documention you create and explains the functionality of the circuit.
\end{itemize}
\textbf{Grading Criteria}
\begin{itemize}
  \item clear and thorough documentation that includes all prescribed components
  \item clear layout of the digital logic design, including appropriate naming conventions
  \item an appropriate problem that illustrates your skills
  \item good use of digital logic principles and components
  \item grammar and spelling
  \item working product
\end{itemize}
\end{document}
